First Order Alternation

### 2019-02-03 quantifiers/body.tex
ok

\paragraph{\em Our Contribution}
We extend the alternating automata model to infinite data alphabets,
by defining a model of computation in which all boolean operations,
including complementation, can be done in linear time.  The control
states are given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the
input consists of an event $a$ from a finite alphabet and a tuple of
data variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
transitions are of the form $q(y_1,\ldots,y_k)
\arrow{a(x_1,\ldots,x_n)}{} \phi(x_1, \ldots, x_n, y_1, \ldots, y_k)$,
where $\phi$ is a formula in the first-order theory of the data
domain. In this model, the arguments of a predicate atom
$q(y_1,\ldots,y_k)$ represent the values of the \emph{internal
  variables} associated with the state. Together with the input values
$x_1,\ldots,x_n$, these values define the next configurations, but
remain invisible in the input sequence.

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{tabular}{ccc}
\begin{minipage}{3cm}
\input ta.pdf_t
\end{minipage} 
&&
\begin{minipage}{5cm}
\small{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]}
\end{minipage} \\
&& 
\begin{minipage}{5cm}
{\small\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{tabular}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered states to be propositional variables, as in the classical
textbook alternating automata model, and all variables of the
automaton to be observable in the input. The model in this paper
overcomes this latter restriction by allowing for internal variables,
whose variables are not visible in the language.

This solves an older language inclusion problem
$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
automata with data variables, whose languages are alternating
sequences of input events and variable valuations
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check. Using
first-order alternating automata allows to bypass this limitation of
our previous work.

The work probably closest to the one reported here concerns the model
of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD},
applied to the verification of parameterized concurrent programs with
shared memory. In this model, the alphabet consists of pairs of
program statements and thread identifiers, thus being infinite because
the number of threads is unbounded. Because thread identifiers can
only be compared for (dis-)equality, the data theory in PA is the
theory of equality. Even with this simplification, the emptiness
problem is undecidable when either the predicates have arity greater
than one \cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of the
contributions of the work reported in this paper.

>>>>

\paragraph{\em Our Contribution}
We extend alternating automata to infinite data alphabets, by defining
a model of computation in which all boolean operations, including
complementation, can be done in linear time.  The control states are
given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the input
consists of an event $a$ from a finite alphabet and a tuple of data
variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
transitions are of the form $q(y_1,\ldots,y_k)
\arrow{a(x_1,\ldots,x_n)}{} \phi(x_1, \ldots, x_n, y_1, \ldots, y_k)$,
where $\phi$ is a formula in the first-order theory of the data
domain. In this model, the arguments of a predicate atom
$q(y_1,\ldots,y_k)$ represent the values of the \emph{internal
  variables} associated with the state. Together with the input values
$x_1,\ldots,x_n$, these values define the next configurations, but
remain invisible in the input sequence.

@@

\begin{figure}[htb]
\vspace*{-\baselineskip}
\begin{center}
\begin{minipage}{5cm}
\input ta.pdf_t
\end{minipage} 
\begin{minipage}{5cm}
\footnotesize{\[A \left\{\begin{array}{rcl}
s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
\end{array}\right.\] 
\[B \left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
\end{array}\right.\]
\[\overline{B}\left\{\begin{array}{rcl}
q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
\end{array}\right.\]}
\end{minipage}
\end{center}
\vspace*{-\baselineskip}
\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
\label{fig:ta}
\vspace*{-\baselineskip}
\end{figure}

@@

\paragraph{\em Related Work} 
The first order alternating automata model presented in this paper
extends the alternating automata with variables ranging over infinite
data considered in \cite{IosifXu18}. There all variables were required
to be observable in the input. We overcome this restriction by
allowing internal (invisible) variables. Another closely related work
\cite{IosifRV16} considers an inclusion between an asynchronous
product of automata $A_1 \times \ldots \times A_n$, extended with data
variables, and a monitor automaton $B$. The semi-algorithm defined
here relies on the assumption that all variables of the observer $B$
must be declared in the automata $A_1, \ldots, A_n$ under check. First
order alternating automata bypass this limitation, since the emptiness
checking semi-algorithm handles invisible variables.

The work probably closest to the one reported here concerns the model
of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used
in the verification of parameterized concurrent programs with shared
memory. In this model, the alphabet consists of pairs of program
statements and thread identifiers, thus being infinite because the
number of threads is unbounded. Because thread identifiers can only be
compared for (dis-)equality, the data theory in PA is the theory of
equality. Even with this simplification, the emptiness problem is
undecidable when either the predicates have arity greater than one
\cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of our
contributions.
